A statistical assessment of influenza intensity thresholds from the moving epidemic and WHO methods

### 2024-10-08 Draft/Draft_MEM.tex
updating manuscript, R2.

In the PISA guidelines \citep{WHO2017}, two statistical methods were recommended for influenza intensity thresholding. These are the \textit{moving epidemic method} (MEM; \citealt{Vega2015}) and the \textit{WHO method} \citep{WHO2014}, also referred to as the \textit{average curves method}. The former has also been recommended by ECDC (e.g., \citealt{ECDC2017}) and is used by the US CDC \citep{Biggerstaff2017}. As will be detailed in Section \ref{sec:definitions}, the WHO and MEM approaches bear important similarities and can be seen as variations of the same general approach. The two methods have been adopted by many public health agencies (see Section \ref{sec:review}). Their statistical properties, however, have not yet been studied. In the present work, we aim to close this gap and derive some recommendations for the computation of influenza intensity thresholds. We study the MEM and WHO methods and suggest extensions to account for small sample sizes and secular trends. We obtain some analytical results on their properties, complemented by simulation experiments based on re-sampling of routine surveillance data from France and the US. These indicate that using the default settings, both the MEM and the WHO method tend to produce too low thresholds. Consequently, the number of season peaks classified as high or very high intensity is higher than intended. The best-behaved thresholds are achieved by combining characteristics of the MEM and WHO methods, along with a correction for small sample sizes. Even this configuration, however, is characterized by modest sensitivity and positive predictive values unless data on many historical seasons are available. This concerns especially thresholds for very high intensity, which should thus be interpreted with care.

The remainder of the article is structured as follows. In Section \ref{sec:desirable_properties} we list desirable properties of a thresholding procedure, which will guide the discussion in the following. Section \ref{sec:definitions} provides definitions of the MEM and WHO methods, highlighting three differences between these otherwise similar approaches. Section \ref{sec:review} consists of an overview of published applications of the MEM and WHO methods, as well as some background on related disease monitoring methods. In Section \ref{sec:formalization_extension} we re-state the thresholding task as a statistical prediction problem and suggest some extensions to the methodology. This is followed by an examination of statistical properties of the methods in Section \ref{sec:analytical_results}. In Section \ref{sec:simulation} we conduct the simulation study before concluding with a discussion in Section \ref{sec:discussion}. Here we also provide practical recommendations on the different implementation choices.

@@

\begin{description}
\item \textbf{Calibration.} As stated in Section \ref{sec:introduction}, thresholds are defined in a statistical manner -- the respective thresholds should be exceeded by 60\%, 10\% and 2.5\% of season peaks. Threshold setting thus corresponds to determining predictive quantiles for future season peak values, and a major requirement is that these are \textit{calibrated}. We call a thresholding method calibrated if in the long run, the intended fractions of seasons are classified as low, medium, high and very high. If, for instance, a method flags new season peaks as very high intensity in considerably more than 2.5\% of the cases, this will hamper the usefulness of thresholds.
\item \textbf{Sensitivity and specificity.} In addition to classifying the intended fraction of seasons into the various categories, the individual classifications should be correct. Assuming there is an underlying stationary distribution of season peaks, a peak which is truly among the 2.5\%/10\%/60\% highest peaks should also be flagged as such (\textit{sensitivity}). Conversely, a peak which is not actually among the 2.5\%/10\%/60\% highest ones should not exceed the respective threshold (\textit{specificity}). These aspects, though, can only be assessed in theory and simulation studies, where the true intensity status can be determined based on the assumed distribution of peaks. In real-world applications, there is no ``gold standard'' to determine the true intensity level.
\item \textbf{Stability.} Estimated thresholds should not be overly variable. Ideally they would not fluctuate strongly around the corresponding quantiles of the true distribution of season peaks.
\item \textbf{Simplicity.} To ensure broad practical applicability, understanding the thresholding method and its parameters should not require advanced statistical training.
\item \textbf{Ease of practical application.} Methods should be straightforward to apply using well-documented and, ideally, open-source software packages.
\end{description}

\section{Definition of the MEM and WHO methods}
\label{sec:definitions}

While framed slightly differently in the respective documentations, the MEM and WHO methods are two special cases of the same general approach. We assume that thresholds are based on weekly values of a transmissibility or severity indicator (see Section \ref{sec:introduction}) from $m$ past seasons and applied to the ($m + 1$)-th season.  We denote the value for season $j$ and week $k$ by $x_{j, k}$, with more specific notation introduced below. Implicitly it is assumed that each season consists of just one wave without multiple peaks separated by longer time spans. This is typically the case in temperate, but not necessarily in tropical regions. Thresholds are then obtained via the following steps.

\begin{enumerate}
\item[(a)] \textbf{Smoothing of historical data} (optional): apply an $l$-week moving average to all historical seasons. If data are smoothed, we denote by $x^\text{raw}_{j, k}$ the raw observation from season $j = 1, \dots, m$, week $k = 1, \dots, 52$, and by
$$
x^\text{smo}_{j, k} = \frac{1}{l} \sum_{d = 0}^{l - 1} x^\text{raw}_{j, k - d}, \ \ k = l, \dots, 52
$$ the smoothed version. In the remainder of this description we denote whichever of $x^\text{raw}_{j, k}$ and $x^\text{smo}_{j, k}$ is used to compute thresholds by $x_{j, k}$.
\item[(b)] \textbf{Sorting:} Within each historical season $j = 1, \dots, m$ order all observations $x_{j, k}$ in decreasing order, denoting the $i$-th largest observation from season $j$ by $x^{(i)}_j$. More generally we will use the notation $^{(\cdot)}$ to denote an ordering of values, e.g., denoting by $x^{\text{raw}, (i)}_j$ the $i$-th largest unsmoothed incidence value from season $j$.
\item[(c)] \textbf{Selection of reference set:} Select the $n$ largest observations from each of the $m$ past seasons to construct a reference set $\mathcal{X} = \{x_j^{(i)}: j = 1, \dots, m; i = 1, \dots, n\}$.
\item[(d)] \textbf{Data transformation:} Apply a monotonically increasing transformation $y_j^{(i)} = f(x_j^{(i)})$ to all members of the reference set $\mathcal{X}$ to obtain a reference set $\mathcal{Y}$ of transformed values.
\item[(e)] \textbf{Fitting a normal distribution:} Assume that the transformed values in $\mathcal{Y}$ come from a normal distribution and compute estimates of its mean and variance,
\begin{align}
\begin{split}
\bar{y} & = \frac{\sum_{j = 1}^m\sum_{i = 1}^n y_j^{(i)}}{n\times m}\\
\sd^2 & = \frac{\sum_{j = 1}^m\sum_{i = 1}^n \left(y_j^{(i)}  - \bar{y}\right)^2}{n\times m - 1}.
\end{split}\label{eq:moments}
\end{align}
\item[(f)] \textbf{Computation of thresholds:} Define intensity thresholds on the transformed scale as quantiles of the normal distribution N$(\bar{y}, \sd^2)$, i.e.\ compute
\begin{equation}
\hat{q}_{Y, \alpha} = \bar{y} + z_\alpha \sd, \label{eq:def_q}
\end{equation}
with $z_\alpha$ the $\alpha$ quantile of the standard normal distribution. The $\hat{q}_{Y, \alpha}$ can be seen as estimates of quantiles $q_{Y, \alpha}$ of an underlying distribution of peak values. The default choices are
\begin{itemize}
\item[(i)] the 40th percentile $\hat{q}_{Y, 0.4} = \bar{y} - 0.25 \sd$ as the threshold for medium intensity;
\item[(ii)] the 90th percentile $\hat{q}_{Y, 0.9} = \bar{y} + 1.28 \sd$ as the threshold for high intensity;
\item[(iii)] the 97.5th percentile $\hat{q}_{Y, 0.975} = \bar{y} + 1.96\sd$ as the threshold for very high intensity.
\end{itemize}
As we will detail in Section \ref{subsec:reformulation}, we will moreover consider an alternative formulation based on the $t$-distribution,
\begin{equation}
\hat{q}_{Y, \alpha} = \bar{y} + t_{m\times n - 1, \alpha} \times \sqrt{1 + \frac{1}{m\times n}} \times s,
\label{eq:q_Y_t}
\end{equation}
with $t_{m\times n - 1, \alpha}$ the $\alpha$ quantile of the $t$ distribution with $m\times n - 1$ degrees of freedom.
\item[(g)] \textbf{Transformation of thresholds to the original scale:} Obtain thresholds on the original scale by applying the inverse transformation, i.e. for $\alpha = 0.4, 0.9, 0.975$ set% $\hat{q}_{X, 0.4} = f^{-1}(\hat{q}_{Y, 0.4})$, $\hat{q}_{X, 0.9} = f^{-1}(\hat{q}_{Y, 0.9})$, $\hat{q}_{X, 0.975} = f^{-1}(\hat{q}_{Y, 0.975})$.
$$
\hat{q}_{X, \alpha} = f^{-1}(\hat{q}_{Y, \alpha}).
$$
\item[(h)] \textbf{Application of thresholds:} The thresholds are applied to classify the peak value of the $(m + 1)$-th season. Depending on the exact specification, thresholds can be applied either to the raw peak value $x_{m + 1}^{\text{raw}, (1)}$ or the smoothed version $x_{m + 1}^{\text{smo}, (1)}$.
\end{enumerate}

\noindent The MEM and WHO methods are special cases of this procedure, see also Table \ref{tab:differences} and the graphical illustration in Figure \ref{fig:illustration}. In the MEM, no smoothing is applied and the default transformation $f$ is the natural logarithm.  \cite{Vega2015} recommend using $5 \leq m \leq 10$ seasons to ensure a recent data basis. The number of observations included per season is set to $n = 30/m$, rounded to the nearest integer (with a minimum value of $n = 1$). The total number $m \times n$ of historical observations is thus kept approximately fixed. This corresponds to the description by \cite{Vega2015} and the default settings of the R package \texttt{mem} \citep{Lozano2020}. The package, however, permits the user to choose $n$, $m$, and $f$, see Supplement B.1. We note that while \texttt{mem} does not currently allow for data smoothing using a moving average, some alternatives are available. Thresholds based on a $t$-distribution as in equation \eqref{eq:q_Y_t} were not implemented when the computations for this paper were run, and an independent custom function was used instead. In the meantime, however, this feature has been added to the development version of the \texttt{mem} package, see Supplement B.1.

@@

\begin{figure}
\begin{center}
\includegraphics[width = 0.78\textwidth]{figure/illustration_mem_who.pdf}
\end{center}
\caption{Illustration of the MEM and WHO methods with their respective default settings, computed using $m = 6$ historical seasons (solid / dashed lines depending on whether smoothing is applied). These thresholds are applied to a seventh season (dotted lines, peak highlighted by a cross). Here, both methods classify the new season peak as medium intensity. Observations included in the reference set are highlighted by dots (with $n = 30/6 = 5$ for MEM, $n = 1$ smoothed values for WHO). The distribution fitted to the reference set is shown as a vertical density plot. Note that for the MEM this is a log-normal as the normal distribution is fitted to log-transformed values. The data used for illustration correspond to the years 2012--2019 from the French region of Grand Est, see Section \ref{subsec:data} and Figure \ref{fig:data}.}
\label{fig:illustration}
\end{figure}

@@

The large number of entries from the years 2019 and 2020 indicates that the MEM has quickly become a standard approach to set intensity thresholds for influenza and other respiratory diseases. The articles come from numerous countries and in many cases have been co-authored by representatives of national or regional public health agencies. In most analyses, threshold levels at the 40th, 90th and 97.5th percentile as in Section \ref{sec:definitions} are used. Variability with respect to the number $m$ of historical seasons included is considerable, with a range from 3 to 16. Consequently, the number $n$ of observations included per season ranges from two to ten. Only one paper \citep{Dahlgren2022} indicated a modification of the default setting $n = 30/m$  and used $n = 3$ with $m = 7$.

@@

\subsection{Related disease monioring methods}
\label{sec:related_literature}

There is a vast literature on monitoring tasks in infectious disease epidemiology and public health. Well-known approaches include Shewhart, cumulative sum (CUSUM, \citealt{Hoehle2008}) and exponentially weighted moving average (EWMA, \citealt{Steiner2010}) control charts. Time series methodology can likewise be employed, see e.g., \cite{Reis2003} for an application of ARIMA models. For a comprehensive account we refer the reader to the review aricles by \cite{Allevius2020}, \cite{Rigdon2015} and \cite{Unkel2012}. In the following we focus on two widely used approaches with close links to the MEM and WHO methods.

A simple Shewhart-type method is used by the US CDC as part of the \textit{EARS} system (Early Aberration Reporting, \citealt{Hutwagner2003}). Denote by $y_1, \dots, y_m$ the $m$ most recent values of a daily epidemiological indicator. An upper alarm threshold for a new observation $y_{m + 1}$ is given by
$$
U = \bar{y} + 3 \times s,
$$ 
where, similarly to equation \eqref{eq:moments} we set $\bar{y} = 1/m \times \sum_{j = 1}^m y_j$ and $s^2 = 1/(m - 1) \times \sum_{j = 1}^m (y_j - \bar{y})^2$. This closely resembles the WHO method, i.e., the procedure from Section \ref{sec:definitions} with $n = 1$ and $f$ the identity function. The multiplier 3 corresponds to $\alpha = 0.9987$. This approach was conceived as a ``drop -in'' surveillance method, i.e., a method to be used in the short term and based on little data. In practice, $m = 7$ observations are used to compute thresholds, which is similar to the typical number of historical seasons used in intensity thresholding.

@@

We now re-state the purpose and assumptions of the WHO and moving epidemic methods in statistical terms, denoting random variables by capital letters (e.g., $X_j^{(i)}$) and their realizations by lower-case letters (e.g., $x_j^{(i)}$). The goal of the thresholding exercise is to use data from seasons 1 through $m$ to obtain values $\hat{q}_{Y, \alpha}$ which shall be exceeded by a new transformed peak value $Y_{m + 1}^{(1)}$ with probabilities $(1 - \alpha) \in \{0.6, 0.1, 0.025\}$, respectively. The $\hat{q}_{Y, \alpha}$ can thus be seen as \textit{predictive quantiles} at levels $\alpha \in \{0.4, 0.9, 0.975\}$. Prediction of new realizations based on parametric assumptions and a set of previous observations is a well-studied problem; see e.g., \cite{Millard2013} on applications in environmental studies. We note that in \cite{WHO2014} and \cite{Vega2015} the thresholds are refered to as the upper ends of one-sided confidence intervals. This, however, is imprecise terminology as in the computations the standard deviation $\sd$ rather than the standard error $\sd/\sqrt{nm}$ is used (see documentation of the \texttt{mem} package and \citealt{WHO2014}, p.69).

@@

As mentioned in Secion \ref{sec:related_literature}, regression approaches are a natural extension of moment-based techniques like the WHO and moving epidemic methods. We illustrate this with a suggestion to account for secular trends in threshold setting, thus relaxing assumption (HB). Secular trends are the reason why for the MEM it is recommend to use no more than $m = 10$ historical seasons, with \citeauthor{Vega2013} (\citeyear{Vega2013}, p.556) cautioning that ``more than ten seasons may further increase accuracy but make a model susceptible to biases from secular trends.'' An obvious model accouning for a trend is
\begin{equation}
Y_j^{(1)} = \beta_0 + \beta_1 \times j + \varepsilon_j, \ \ \ \ \varepsilon_j \stackrel{\text{i.i.d.}}{\sim} \text{N}(0, \sigma^2_\varepsilon).\label{eq:trend}
\end{equation}
Parameter estimates $\hat{\alpha}$ and $\hat{\beta}$ can be computed using peak values $y_1^{(1)}, \dots, y_m^{(1)}$ from the last $m$ seasons and the ordinary least squares method. An unbiased estimator of ${\sigma}^2_\varepsilon$ is
$$
\hat{\sigma}^2_\varepsilon = \frac{1}{m - 2} \sum_{j = 1}^m (y^{(1)}_i - \beta_0 - \beta_1 \times j)^2.
$$
Predictive quantiles at levels $\alpha = 0.4, 0.9, 0.975$, and thus intensity thresholds, can be obtained using standard methods implemented in software packages such as R. As we show in Supplement B.2, the generalization of equation \eqref{eq:q_Y_t} resulting for model \eqref{eq:trend} is of a simple form. A graphical illustration of the resulting time-varying thresholds can be found in Figure \ref{fig:illustration_trend}. This approach could easily be extended to accommodate $n > 1$ values per historical season. However, as we will argue in Section \ref{subsec:choice_n}, this will lead to biased thresholds, unless a more flexible model formulation accounting for relevant mean and correlation structures is adopted. We therefore only consider $n = 1$.

@@

In this section we provide some statistical reasoning on the consequences of various implementation choices within the thresholding framework outlined in Section \ref{sec:definitions}. Application-focused readers may find the simulation-based illustration of our findings in Section \ref{sec:simulation} more vivid.

@@

\noindent \textbf{Thresholds and threshold exceedance.} Figure \ref{fig:trend} shows that ignoring secular trends leads to mis-calibrated thresholds, with stronger miscalibration for larger $m$. This is expected as seasons far in the past in expectation differ more strongly from the upcoming $(m + 1)$-th one. Thresholds corrected for secular trends using model \eqref{eq:trend} are well-calibrated. However, due to the addition of a parameter, the threshold values get considerably more variable. The corresponding display for $r = \pm 7\%$ is available in Supplementary Figure S5 and shows the same qualitative patterns.

@@

\noindent \textbf{Limitations.} The present study focuses on the statistical properties of thresholding methods, but we emphasize that intensity classification also requires in-depth domain knowledge. Indeed, some experts we consulted stated that in their institutions weekly classifications for international reports were occasionally overruled manually if there was reason to believe that current data were biased in a specific way. This illustrates that while statistical methods should of course be well-behaved, they are only one element in a more complex assessment process.

>>>>

In the PISA guidelines \citep{WHO2017}, two statistical methods were recommended for influenza intensity thresholding. These are the \textit{moving epidemic method} (MEM; \citealt{Vega2015}) and the \textit{WHO method} \citep{WHO2014}, also referred to as the \textit{average curves method}. The former has also been recommended by ECDC (e.g., \citealt{ECDC2017}) and is used by the US CDC \citep{Biggerstaff2017}. As will be detailed in Section \ref{sec:mem_and_who}, the WHO and MEM approaches bear important similarities and can be seen as variations of the same general approach. The two methods have been adopted by many public health agencies (see Section \ref{sec:review}). Their statistical properties, however, have not yet been studied. In the present work, we aim to close this gap and derive some recommendations for the computation of influenza intensity thresholds. We study the MEM and WHO methods and suggest extensions to account for small sample sizes and secular trends. We obtain some analytical results on their properties, complemented by simulation experiments based on re-sampling of routine surveillance data from France and the US. These indicate that using the default settings, both the MEM and the WHO method tend to produce too low thresholds. Consequently, the number of season peaks classified as high or very high intensity is higher than intended. The best-behaved thresholds are achieved by combining characteristics of the MEM and WHO methods, along with a correction for small sample sizes. Even this configuration, however, is characterized by modest sensitivity and positive predictive values unless data on many historical seasons are available. This concerns especially thresholds for very high intensity, which should thus be interpreted with care.

The remainder of the article is structured as follows. In Section \ref{sec:desirable_properties} we list desirable properties of a thresholding procedure, which will guide the discussion in the following. Section \ref{sec:mem_and_who} provides definitions of the MEM and WHO methods, highlighting three differences between these otherwise similar approaches. Section \ref{sec:review} consists of an overview of published applications of the MEM and WHO methods, as well as some background on related disease monitoring methods. In Section \ref{sec:formalization_extension} we re-state the thresholding task as a statistical prediction problem and suggest some extensions to the methodology. This is followed by an examination of statistical properties of the methods in Section \ref{sec:analytical_results}. In Section \ref{sec:simulation} we conduct the simulation study before concluding with a discussion in Section \ref{sec:discussion}. Here we also provide practical recommendations on the different implementation choices.

@@

\begin{description}
\item \textbf{Calibration.} As stated in Section \ref{sec:introduction}, thresholds are defined in a statistical manner -- the respective thresholds should on average be exceeded by 60\%, 10\% and 2.5\% of season peaks. Threshold setting thus corresponds to determining predictive quantiles for future season peak values, and a major requirement is that these are \textit{calibrated}. We call a thresholding method calibrated if in the long run, the intended fractions of seasons are classified as low, medium, high and very high. If, for instance, a method flags new season peaks as very high intensity in considerably more than 2.5\% of the cases, this will hamper the usefulness of thresholds.
\item \textbf{Sensitivity and specificity.} In addition to classifying the intended fraction of seasons into the various categories, the individual classifications should be correct. Assuming there is an underlying stationary distribution of season peaks, a peak which is truly among the 2.5\%/10\%/60\% highest peaks should also be flagged as such (\textit{sensitivity}). Conversely, a peak which is not actually among the 2.5\%/10\%/60\% highest ones should not exceed the respective threshold (\textit{specificity}). These aspects, though, can only be assessed in theory and simulation studies, where the true intensity status can be determined based on the assumed distribution of peaks. In real-world applications, there is no ``gold standard'' to determine the true intensity level.
\item \textbf{Stability.} Estimated thresholds should not be overly variable. Ideally they would not fluctuate strongly around the corresponding quantiles of the true distribution of season peaks.
\item \textbf{Simplicity.} To ensure broad practical applicability, understanding the thresholding method and its parameters should not require advanced statistical training.
\item \textbf{Ease of practical application.} Methods should be straightforward to apply using well-documented and, ideally, open-source software packages.
\end{description}

\section{The MEM and WHO methods}
\label{sec:mem_and_who}

\subsection{An overarching definition}
\label{subsec:definitions}

While framed slightly differently in the respective documentations, the MEM and WHO methods are two special cases of the same general approach. We assume that thresholds are based on weekly values of a transmissibility or severity indicator (see Section \ref{sec:introduction}) from $m$ past seasons and applied to the ($m + 1$)-th season.  We denote the indicator value for season $j$ and week $k$ by $x_{j, k}$, with more specific notation introduced below. Implicitly it is assumed that each season consists of just one wave without multiple peaks separated by longer time spans.

\begin{enumerate}
\item[(a)] \textbf{Smoothing of historical data} (optional): apply an $l$-week moving average to all historical seasons. If data are smoothed, we denote by $x^\text{raw}_{j, k}$ the raw observation from season $j = 1, \dots, m$, week $k = 1, \dots, 52$, and by
$$
x^\text{smo}_{j, k} = \frac{1}{l} \sum_{d = 0}^{l - 1} x^\text{raw}_{j, k - d}, \ \ k = l, \dots, 52
$$ the smoothed version. In the remainder of this description we denote whichever of $x^\text{raw}_{j, k}$ and $x^\text{smo}_{j, k}$ is used to compute thresholds by $x_{j, k}$.
\item[(b)] \textbf{Sorting:} Within each historical season $j = 1, \dots, m$ order all observations $x_{j, k}$ in decreasing order, denoting the $i$-th largest observation from season $j$ by $x^{(i)}_j$. More generally we will use the notation $^{(\cdot)}$ to denote an ordering of values, e.g., denoting by $x^{\text{raw}, (i)}_j$ the $i$-th largest unsmoothed incidence value from season $j$.
\item[(c)] \textbf{Selection of reference set:} Select the $n$ largest observations from each of the $m$ past seasons to construct a reference set $\mathcal{X} = \{x_j^{(i)}: j = 1, \dots, m; i = 1, \dots, n\}$.
\item[(d)] \textbf{Data transformation:} Apply a monotonically increasing transformation $y_j^{(i)} = f(x_j^{(i)})$ to all members of the reference set $\mathcal{X}$ to obtain a reference set $\mathcal{Y}$ of transformed values.
\item[(e)] \textbf{Fitting a normal distribution:} Assume that the transformed values in $\mathcal{Y}$ are independent and identically distributed (i.i.d.) draws from a normal distribution and compute estimates of its mean and variance,
\begin{align}
\begin{split}
\bar{y} & = \frac{\sum_{j = 1}^m\sum_{i = 1}^n y_j^{(i)}}{n\times m}\\
\sd^2 & = \frac{\sum_{j = 1}^m\sum_{i = 1}^n \left(y_j^{(i)}  - \bar{y}\right)^2}{n\times m - 1}.
\end{split}\label{eq:moments}
\end{align}
\item[(f)] \textbf{Computation of thresholds:} Define intensity thresholds on the transformed scale as quantiles of the normal distribution N$(\bar{y}, \sd^2)$, i.e.\ compute
\begin{equation}
\hat{q}_{Y, \alpha} = \bar{y} + z_\alpha \sd, \label{eq:def_q}
\end{equation}
with $z_\alpha$ the $\alpha$ quantile of the standard normal distribution. The $\hat{q}_{Y, \alpha}$ can be seen as estimates of quantiles $q_{Y, \alpha}$ of an underlying distribution of peak values. The default choices are
\begin{itemize}
\item[(i)] the 40th percentile $\hat{q}_{Y, 0.4} = \bar{y} - 0.25 \sd$ as the threshold for medium intensity;
\item[(ii)] the 90th percentile $\hat{q}_{Y, 0.9} = \bar{y} + 1.28 \sd$ as the threshold for high intensity;
\item[(iii)] the 97.5th percentile $\hat{q}_{Y, 0.975} = \bar{y} + 1.96\sd$ as the threshold for very high intensity.
\end{itemize}
As we will detail in Section \ref{subsec:reformulation}, we will moreover consider an alternative formulation based on the $t$-distribution,
\begin{equation}
\hat{q}_{Y, \alpha} = \bar{y} + t_{m\times n - 1, \alpha} \times \sqrt{1 + \frac{1}{m\times n}} \times s,
\label{eq:q_Y_t}
\end{equation}
with $t_{m\times n - 1, \alpha}$ the $\alpha$ quantile of the $t$ distribution with $m\times n - 1$ degrees of freedom.
\item[(g)] \textbf{Transformation of thresholds to the original scale:} Obtain thresholds on the original scale by applying the inverse transformation, i.e. for $\alpha = 0.4, 0.9, 0.975$ set% $\hat{q}_{X, 0.4} = f^{-1}(\hat{q}_{Y, 0.4})$, $\hat{q}_{X, 0.9} = f^{-1}(\hat{q}_{Y, 0.9})$, $\hat{q}_{X, 0.975} = f^{-1}(\hat{q}_{Y, 0.975})$.
$$
\hat{q}_{X, \alpha} = f^{-1}(\hat{q}_{Y, \alpha}).
$$
\item[(h)] \textbf{Application of thresholds:} The thresholds are applied to classify the peak value of the $(m + 1)$-th season. Depending on the exact specification, thresholds can be applied either to the raw peak value $x_{m + 1}^{\text{raw}, (1)}$ or the smoothed version $x_{m + 1}^{\text{smo}, (1)}$.
\end{enumerate}

\subsection{Specifics and implementations}

The MEM and WHO methods are special cases of this procedure, see also Table \ref{tab:differences} and the graphical illustration in Figure \ref{fig:illustration}. In the MEM, no smoothing is applied and the default transformation $f$ is the natural logarithm.  \cite{Vega2015} recommend using $5 \leq m \leq 10$ seasons to ensure a recent data basis. The number of observations included per season is set to $n = 30/m$, rounded to the nearest integer (with a minimum value of $n = 1$). The total number $m \times n$ of historical observations is thus kept approximately fixed. This corresponds to the description by \cite{Vega2015} and the default settings of the R package \texttt{mem} \citep{Lozano2020}. The package, however, permits the user to choose $n$, $m$, and $f$, see Supplement B.1. We note that while \texttt{mem} does not currently allow for data smoothing using a moving average, some alternatives are available. Thresholds based on a $t$-distribution as in equation \eqref{eq:q_Y_t} were not implemented when the computations for this paper were run, and an independent custom function was used instead. In the meantime, however, this feature has been added to the development version of the \texttt{mem} package, see Supplement B.1.

@@

\begin{figure}
\begin{center}
\includegraphics[width = 0.78\textwidth]{figure/illustration_mem_who.pdf}
\end{center}
\caption{Illustration of the MEM and WHO methods with their respective default settings, computed using $m = 6$ historical seasons (solid / dashed lines depending on whether smoothing is applied). These thresholds are applied to a seventh season (dotted lines, peak highlighted by a cross). Here, both methods classify the new season peak as medium intensity. Identical thresholds are also displayed for past seasons, but greyed out as the real-time assessment of these seasons would have been based on different thresholds computed from the respective preceding data. Observations included in the reference set are highlighted by dots (with $n = 30/6 = 5$ for MEM, $n = 1$ smoothed values for WHO). The distribution fitted to the reference set is shown as a vertical density plot. Note that for the MEM this is a log-normal as the normal distribution is fitted to log-transformed values. The data used for illustration correspond to the years 2012--2019 from the French region of Grand Est, see Section \ref{subsec:data} and Figure \ref{fig:data}.}
\label{fig:illustration}
\end{figure}

@@

The large number of entries from the years 2019 and 2020 indicates that the MEM has quickly become a standard approach to set intensity thresholds for influenza and other respiratory diseases. The articles come from numerous countries and in many cases have been co-authored by representatives of national or regional public health agencies. In most analyses, threshold levels at the 40th, 90th and 97.5th percentile as in Section \ref{subsec:definitions} are used. Variability with respect to the number $m$ of historical seasons included is considerable, with a range from 3 to 16. Consequently, the number $n$ of observations included per season ranges from two to ten. Only one paper \citep{Dahlgren2022} indicated a modification of the default setting $n = 30/m$  and used $n = 3$ with $m = 7$.

@@

\subsection{Related disease monitoring methods}
\label{sec:related_literature}

There is a vast literature on monitoring tasks in infectious disease epidemiology and public health. Similarly to the peak classification problem we address, the aim is to detect and characterize unusual events, and the underlying mathematics are often similar (see also Section \ref{sec:formalization_extension}). Well-known approaches include Shewhart, cumulative sum (CUSUM, \citealt{Hoehle2008}) and exponentially weighted moving average (EWMA, \citealt{Steiner2010}) control charts. Time series methodology can likewise be employed, see e.g., \cite{Reis2003} for an application of ARIMA models. For a comprehensive account we refer the reader to the review aricles by \cite{Allevius2020}, \cite{Rigdon2015} and \cite{Unkel2012}. In the following we focus on two widely used approaches with close links to the MEM and WHO methods.

A simple Shewhart-type method is used by the US CDC as part of the \textit{EARS} system (Early Aberration Reporting, \citealt{Hutwagner2003}). Denote by $y_1, \dots, y_m$ the $m$ most recent values of a daily epidemiological indicator. An upper alarm threshold for a new observation $y_{m + 1}$ is given by
$$
U = \bar{y} + 3 \times s,
$$ 
where, similarly to equation \eqref{eq:moments} we set $\bar{y} = 1/m \times \sum_{j = 1}^m y_j$ and $s^2 = 1/(m - 1) \times \sum_{j = 1}^m (y_j - \bar{y})^2$. This closely resembles the WHO method, i.e., the procedure from Section \ref{subsec:definitions} with $n = 1$ and $f$ the identity function. The multiplier 3 corresponds to $\alpha = 0.9987$. This approach was conceived as a ``drop-in'' surveillance method, i.e., a method to be used in the short term and based on little data. In practice, $m = 7$ observations are used to compute thresholds, which is similar to the typical number of historical seasons used in intensity thresholding.

@@

We now re-state the purpose and assumptions of the WHO and moving epidemic methods in statistical terms, denoting random variables by capital letters (e.g., $X_j^{(i)}$) and their realizations by lower-case letters (e.g., $x_j^{(i)}$). The goal of the thresholding exercise is to use data from seasons 1 through $m$ to obtain values $\hat{q}_{Y, \alpha}$ which shall be exceeded by a new transformed peak value $Y_{m + 1}^{(1)}$ with probabilities $(1 - \alpha) \in \{0.6, 0.1, 0.025\}$, respectively. The $\hat{q}_{Y, \alpha}$ can thus be seen as \textit{predictive quantiles} at levels $\alpha \in \{0.4, 0.9, 0.975\}$. Prediction of new realizations based on parametric assumptions and a set of previous observations is a well-studied problem; see e.g., \cite{Millard2013} on applications in environmental studies. We note that in \cite{WHO2014} and \cite{Vega2015} the thresholds are refered to as the upper ends of one-sided confidence intervals. This, however, is imprecise terminology as in the computations the standard deviation $s$ (of the reference observations) rather than the standard error $s/\sqrt{nm}$ (of the sample mean $\bar{y}$) is used. See documentation of the \texttt{mem} package and \citet[p.69]{WHO2014}.

@@

As mentioned in Secion \ref{sec:related_literature}, regression approaches are a natural extension of moment-based techniques like the WHO and moving epidemic methods. We illustrate this with a suggestion to account for secular trends in threshold setting, thus relaxing assumption (HB). Secular trends are the reason why for the MEM it is recommend to use no more than $m = 10$ historical seasons, with \citeauthor{Vega2013} (\citeyear{Vega2013}, p.556) cautioning that ``more than ten seasons may further increase accuracy but make a model susceptible to biases from secular trends.'' A simple model accouning for a trend is
\begin{equation}
Y_j^{(1)} = \beta_0 + \beta_1 \times j + \varepsilon_j, \ \ \ \ \varepsilon_j \stackrel{\text{i.i.d.}}{\sim} \text{N}(0, \sigma^2_\varepsilon).\label{eq:trend}
\end{equation}
Parameter estimates $\hat{\alpha}$ and $\hat{\beta}$ can be computed using peak values $y_1^{(1)}, \dots, y_m^{(1)}$ from the last $m$ seasons and the ordinary least squares method. An unbiased estimator of ${\sigma}^2_\varepsilon$ is
$$
\hat{\sigma}^2_\varepsilon = \frac{1}{m - 2} \sum_{j = 1}^m (y^{(1)}_i - \beta_0 - \beta_1 \times j)^2.
$$
Predictive quantiles at levels $\alpha = 0.4, 0.9, 0.975$, and thus intensity thresholds, can be obtained using standard methods implemented in software packages such as R. As we show in Supplement B.2, the generalization of equation \eqref{eq:q_Y_t} resulting for model \eqref{eq:trend} is of a simple form. A graphical illustration of the resulting time-varying thresholds can be found in Figure \ref{fig:illustration_trend}. This approach could easily be extended to accommodate $n > 1$ values per historical season. However, as we will argue in Section \ref{subsec:choice_n}, this will lead to biased thresholds, unless a more flexible model formulation accounting for relevant mean and correlation structures is adopted. We therefore only consider $n = 1$.

@@

In this section we provide some statistical reasoning on the consequences of various implementation choices within the thresholding framework outlined in Section \ref{subsec:definitions}. Application-focused readers may find the simulation-based illustration of our findings in Section \ref{sec:simulation} more vivid.

@@

\noindent \textbf{Thresholds and threshold exceedance.} Figure \ref{fig:trend} shows that ignoring secular trends leads to miscalibrated thresholds, with stronger miscalibration for larger $m$. This is expected as seasons far in the past in expectation differ more strongly from the upcoming $(m + 1)$-th one. Thresholds corrected for secular trends using model \eqref{eq:trend} are well-calibrated. However, due to the addition of a parameter, the threshold values get considerably more variable. The corresponding display for $r = \pm 7\%$ is available in Supplementary Figure S5 and shows the same qualitative patterns.

@@

\noindent \textbf{Limitations.} The present study focuses on the statistical properties of thresholding methods, but we emphasize that intensity classification also requires in-depth domain knowledge. Indeed, some experts we consulted stated that in their institutions, weekly classifications for international reports were occasionally overruled manually if there was reason to believe that current data were biased in a specific way. This illustrates that while statistical methods should of course be well-behaved, they are only one element in a more complex assessment process.
